{\bf Affine.m} -- {\it Mathematica} package for computations in representation theory of finite-dimensional and affine Lie algebras

### 2011-05-14 doc/paper.tex
Some code updates, paper outline is finished.

The structure of module is determined by the set of weights since the action of generators $E^{\alpha}$ on weight vectors is
\begin{equation}
  \label{eq:5}
  E^{\alpha}\cdot v_{\lambda} \propto v_{\lambda+\alpha}
\end{equation}
It can be encoded by the formal character of module
\begin{equation}
  \label{eq:10}
  \mathrm{ch}V=\sum_{\lambda}m_{\lambda} e^{\lambda}
\end{equation}
Character  $\mathrm{ch}V\in \mathcal{E}$. is the element of algebra $\mathcal{E}$ generated by formal exponents of weights.
It is possible to specialise character by taking its value on some element $\xi$ of $\hf$:

Roots are weights of adjoint representation of $\gf$ on itself. They encode the commutation relations of algebra in the following way. Denote by $\Delta$ the set of roots. For each $\alpha\in \Delta$ there exist root $-\alpha\in \Delta$ and generators $E^{\alpha}, E^{-\alpha}$ such that
\begin{align}
  \label{eq:4}
  &  [H^{i},E^{\alpha}]=\alpha^{i}E^{\alpha} \\
  &\left[E^{\alpha},E^{\beta}\right]=
  \begin{cases} 
    N_{\alpha,\beta} E^{\alpha+\beta}, & \mbox{if}\; \alpha+\beta\in \Delta\\
    \frac{2}{(\alpha,\alpha)} \sum_{i}\alpha^{i} H^{i},&  \mbox{if}\; \alpha=-\beta\\
    0,&\mbox{otherwise}
  \end{cases}
\end{align}

@@

Roots system can be reconstructed from the set of simple roots with the use of Weyl group $W_{\gf}$ which is generated by reflections $\{s_{i}:\hfg^{*}\to\hfg^{*}\}$ corresponding to simple roots $\{\alpha_{i}\}$:
\begin{equation}
  \label{eq:8}
  s_{i}\cdot\lambda=\lambda-\frac{2(\alpha_{i},\lambda)}{(\alpha_{i},\alpha_{i})}\alpha_{i}
\end{equation}
Note that root multiplicity $\mathrm{mult}(\alpha)$ for affine Lie algebra can be greater than one. Weyl vector is denoted by $\rho=\frac{1}{2}\sum_{\alpha\in \Delta^{+}}\alpha$.

@@

\subsection{Highest weight modules}
\label{sec:high-weight-modul}

@@

@@

$\mathrm{Print}\left[B_2\right]$
\begin{lstlisting}
finiteRootSystem[2, 2, {finiteWeight[2, {1, -1}], finiteWeight[2,{0, 1}]}]}
\end{lstlisting}

@@

Semisimple Lie algebras can be created as sums of simple:
\begin{equation*}
  \mathrm{Print}[A_1+A_1]
\end{equation*}
\begin{lstlisting}
  finiteRootSystem
\end{lstlisting}

@@

\section{Examples}
\label{sec:examples}

\section{Conclusion}
\label{sec:conclusion}

>>>>

The structure of module is determined by the set of weights since the action of generators $E^{\alpha}$ on weight vectors is
\begin{equation}
  \label{eq:5}
  E^{\alpha}\cdot v_{\lambda} \propto v_{\lambda+\alpha}
\end{equation}
It can be encoded by the formal character of module
\begin{equation}
  \label{eq:10}
  \mathrm{ch}V=\sum_{\lambda}m_{\lambda} e^{\lambda}
\end{equation}
Character  $\mathrm{ch}V\in \mathcal{E}$ is the element of algebra $\mathcal{E}$ generated by formal exponents of weights.
It is possible to specialise character by taking its value on some element $\xi$ of $\hf$:

Roots are weights of adjoint representation of $\gf$ on itself. They encode the commutation relations of algebra in the following way. Denote by $\Delta$ the set of roots. For each $\alpha\in \Delta$ there exist root $-\alpha\in \Delta$ and generators $E^{\alpha}, E^{-\alpha}$ such that
\begin{align}
  \label{eq:4}
  &  [H^{i},E^{\alpha}]=\alpha^{i}E^{\alpha} \\
  &\left[E^{\alpha},E^{\beta}\right]=
  \begin{cases} 
    N_{\alpha,\beta} E^{\alpha+\beta}, & \mbox{if}\; \alpha+\beta\in \Delta\\
    \frac{2}{(\alpha,\alpha)} \sum_{i}\alpha^{i} H^{i},&  \mbox{if}\; \alpha=-\beta\\
    0,&\mbox{otherwise}
  \end{cases}
\end{align}
Highest weight of adjoint representation.

%%  Given a root system Φ we can always choose (in many ways) a set of positive roots. This is a subset Φ + of Φ such that
%%  
%%      * For each root \alpha\in\Phi exactly one of the roots α, –α is contained in Φ + .
%%      * For any two distinct \alpha, \beta\in \Phi^+ such that α + β is a root, \alpha+\beta\in\Phi^+.
%%  
%%  If a set of positive roots Φ + is chosen, elements of –Φ + are called negative roots.
%%  
%%  An element of Φ + is called a simple root if it cannot be written as the sum of two elements of Φ + . The set Δ of simple roots is a basis of V with the property that every vector in Φ is a linear combination of elements of Δ with all coefficients non-negative, or all coefficients non-positive. For each choice of positive roots, the corresponding set of simple roots is the unique set of roots such that the positive roots are exactly those that can be expressed as a combination of them with non-negative coefficients, and such that these combinations are unique.
%%  
%%  Ordering. Highest root. Rho=sum of fundamental weights, not half-sum of positive roots.

@@

Roots system can be reconstructed from the set of simple roots with the use of Weyl group $W_{\gf}$ which is generated by reflections $\{s_{i}:\hfg^{*}\to\hfg^{*}\}$ corresponding to simple roots $\{\alpha_{i}\}$:
\begin{equation}
  \label{eq:8}
  s_{i}\cdot\lambda=\lambda-\frac{2(\alpha_{i},\lambda)}{(\alpha_{i},\alpha_{i})}\alpha_{i}
\end{equation}
Note that root multiplicity $\mathrm{mult}(\alpha)$ for affine Lie algebra can be greater than one.

@@

Weyl vector is denoted by $\rho=\sum_{i} \omega_{i}$, it is of importance for 
\subsection{Highest weight modules}
\label{sec:high-weight-modul}

@@

%% Define external border of representation

@@

\begin{lstlisting}[mathescape=true]
  $B_2$ == finiteRootSystem[2, 2, {finiteWeight[2, {1, -1}], finiteWeight[2,{0, 1}]}]}
\end{lstlisting}

@@

Semisimple Lie algebras can be created as sums of simple:
\begin{lstlisting}[mathescape=true]
  $A_1$+$A_1$ == finiteRootSystem[2, 2, {finiteWeight[2, {1, 0}], finiteWeight[2, {0, 1}]}]
\end{lstlisting}

Predicate \lstinline{rootSystemQ} checks if the object is root systems of finite or affine type.

List of simple roots is the property of root system so it is accessed as \lstinline{rs[simpleRoots]}.

We have implemented several functions to get major properties of root systems. Weyl vector is given by the function \lstinline{rho[rs_?rootSystemQ]}:
\begin{lstlisting}[label=list:1]
  rho[b2]
  finiteWeight[2, {3/2, 1/2}]
\end{lstlisting}
Positive roots can be constructed with \lstinline{positiveRoots[rs_?rootSystemQ]}. For affine Lie algebra this and related functions return the list up to fixed grade. This grade limit is set through the value of \lstinline{rs[gradeLimit]} which is equal to 10 by default. List of roots (up to \lstinline{gradeLimit}) is returned by \lstinline{roots[rs]}. Cartan matrix and fundamental weights are calculated by the functions \lstinline{cartanMatrix} and \lstinline{fundamentalWeights} correspondingly.

It is possible to specify a weight of Lie algebra by its Dynkin label
\begin{lstlisting}
  weight[b2][1,2,3] == makeFiniteWeight[{2, 1}]
\end{lstlisting}
The function \lstinline{dynkinLabels[rs_?rootSystemQ][wg_?weightQ]} returns Dynkin labels of weight \lstinline{wg} in the root system \lstinline{rs}.

Elements of Weyl group are specified by the set of indices of reflections, so element of Weyl group of algebra $B_{2}$ which is equal to $w=s_{1}s_{2}s_{1}$ is constructed with \lstinline{weylGroupElement[b2][1,2,1]}. Then it can be applied to the weights:
\begin{lstlisting}
  w = weylGroupElement[b2][1,2,1];
  w @ makeFiniteWeight[{1,0}] == makeFiniteWeight[{-1,0}]
\end{lstlisting}

Computation of ShortLex form \cite{casselman1994machine,casselman1995automata} for Weyl group elements can be conveniently implemented through the use of pattern-matching features of {\it Mathematica}. In \cite{KallenShortlex} rewrite rules for simple finite dimensional and affine Lie algebras are presented as {\it Mathematica} patterns. Our presentation of Weyl group elements is compatible with the code of \cite{KallenShortlex}
\begin{lstlisting}[mathescape=true]
  In[1]  = $<<$A3reduce;
           reduce[s[1,2,1,2,1,3,2,1,1]]
  Out[1] = s[2, 3, 2]

  In[2]  = (weylGroupElement[$A_{3}$] @@ reduce[s[1,2,1,2,1,3,2,1,1]]) @ weight[$A_{3}$][-1,-2,-1]
  Out[2] = finiteWeight[4, {-2, 2, 1, -1}]

  In[3]  = dynkinLabels[$A_{3}$][Out[2]]
  Out[3] = {-4, 1, 2}
\end{lstlisting}

\subsection{Formal elements}
\label{sec:formal-elements}

We represent formal characters of representations by special structure \lstinline{formalElements}. This structure is a hash-table implemented with \lstinline{DownValues}. The keys are weights at the exponents and the values are corresponding multiplicities. \lstinline[mathescape=true]!makeFormalElement[{$\gamma_{1},\dots,\gamma_{n}$},{$m_{1},\dots,m_{n}$}]! creates data-structure which represents the element $\sum_{i=1}^{n} m_{i} e^{\gamma_{i}}$ of formal algebra $\mathcal{E}$. The operations of formal algebra $\mathcal{E}$ are implemented for \lstinline{formalElement} data-type: formal elements can be added, multiplied by number or exponent of weight. There is also multiplication of formal elements but no division.
\begin{lstlisting}[mathescape=true]
  In[1]  = makeFormalElement[{makeFiniteWeight[{1,1}],makeFiniteWeight[{0,0}]},{1,2}] *
             (2 * Exp[makeFiniteWeight[{1,0}]] *
             makeFormalElement[{makeFiniteWeight[{1,1}],makeFiniteWeight[{0,0}]},{1,2}]);
  In[2]  = In[1][weights]
  Out[2] = {finiteWeight[2, {1, 0}], finiteWeight[2, {2, 1}], finiteWeight[2, {3, 2}]}

  In[3]  = In[1][multiplicities]
  Out[3]= {8, 8, 2}
\end{lstlisting}

@@

As we have already stated in section \ref{sec:high-weight-modul} there exist two recurrent relations which can be used to calculate weight multiplicities in irreducible modules. Both algorithms proceed in the following way to calculate weight multiplicities:
\begin{enumerate}
\item Create the list of weights in main Weyl chamber by subtraction of all possible combinations of simple roots from the highest weight (e.g. for finite-dimensional algebra subtract $\alpha_{1}$ from $\mu$ while inside $\bar C$, then subtract $\alpha_{2}$ from all the weights already obtained etc).
\item Sort the list of weights by their product with Weyl vector.
\item Use a recurrent formula. If the weight required for recurrent computation is outside the main chamber use Weyl symmetry. 
\end{enumerate}
The difference in performance of algorithms is in the number of previous values required to get the multiplicity of weight under consideration. For recurrent relation \eqref{eq:14} it is constant and equal to the number of elements in Weyl group (if we are far from the boundary of representation diagram). When Freudenthal formula \eqref{eq:15} is used number of previous values grows with the distance from the external border of representation. So Freudenthal formula is faster if the weight is close to the border or the rank of the algebra and the size of Weyl group is large \cite{moody1982fast}. 
But Freudenthal formula is of no use if we are interested in Verma modules.

In the calculation of branching coefficients use of Freudenthal formula requires full construction of formal characters of algebra representation and all the representations of subalgebra. It is impractical when the rank of algebra and subalgebra are big, for example for maximal subalgebra.

Branching algorithm \cite{2010arXiv1007.0318L}
\subsection{Performance testing}
\label{sec:performance-testing}

We have made some experiments with our implementations of Freudenthal formula and formula \eqref{eq:15} and have got the following graphs \ref{freudenthal-racah-times}.

\begin{figure}[h]
  \label{freudenthal-racah-times}
  \noindent\centering{
 %   \includegraphics[width=80mm]{figure1.eps}
  }
  \caption{Freudenthal run time with the number of weights in $\bar C$.}

  \noindent\centering{
 %   \includegraphics[width=80mm]{figure1.eps}
  }
  \caption{Racah run time with the number of weights in $\bar C$.}

  \noindent\centering{
 %   \includegraphics[width=80mm]{figure1.eps}
  }
  \caption{Freudenthal run time with rank of the algebra}

  \noindent\centering{
 %   \includegraphics[width=80mm]{figure1.eps}
  }
  \caption{Racah run time with rank of the algebra}
\end{figure}

Measurements for branching computations for $\hat B_{3}\subset \hat B_{4}$ are shown at the Fig. \ref{fig:branching}.
\begin{figure}[h]
  \label{fig:branching}
  \noindent\centering{
 %   \includegraphics[width=80mm]{figure1.eps}
  }
  \caption{Branching}
\end{figure}

\section{Examples}
\label{sec:examples}
In this section we present some examples of computations available with {\bf Affine.m} with the code required to produce these results.

\section{Conclusion}
\label{sec:conclusion}
We present the package {\bf Affine.m} for computations in representation theory of finite-dimensional and affine Lie algebras. The package is available from {\it github} and is also submitted to the library.
